A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-11 tex/codatadp.tex
dumped email

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$). Symbols starting with an uppercase
letter (e.g.,
$\const{S}$, $\const{SCons}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed compactly as
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$,
which collapses to $\const{f} : \tau$ if $n = 0$.
For a term $t$, the notation $t : \tau$ indicates that it has type $\tau$.
Functions are invoked in the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
%
Finally, the notation $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$
and $\bigwedge_{\,i} \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$.

@@

We fix a signature $\Sig = (\Types, \Funcs)$. The types are partioned into
$\Types = \Data \mathrel{\uplus} \Codata \mathrel{\uplus} \Nondata$, where $\Data$ are the
\emph{datatypes}, $\Codata$ are the \emph{codatatypes}, and $\Nondata$ are the remaining
\emph{ordinary types}. The functions are partitioned into $\Funcs = \Ctr
\mathrel{\uplus} \Sel$, where $\Ctr$ are the \emph{constructors} and $\Sel$ are the
\emph{selectors}; there is no need to consider other function symbols because they can
be abstracted away as variables when combining theories.
%
In an SMT problem, the signature is normally given by specifying the
uninterpreted types in any order, the (co)datatypes with their constructors
and selectors in groups of mutually (co)recursive groups of (co)datatypes, and
finally any other function symbols.

\paragraph{Specifications.}
A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
In the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
through subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}.

@@

For types with several constructors, it is customary to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. However,
it is not necessary to extend the signature:
The discriminator term $\const{d}_{i\negvthinspace j}(x)$ can be seen as an abbreviation for
$x = \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(x), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(x)\bigr)$.
This will simplify the presentation later. % of the decision procedure.

Here are a few examples of legal specifications of (co)datatype families:
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree}_{\,\tau} & \,=\, \const{Node}(\tau,\: \ty{forest}_{\,\tau}) \\[-\jot]
      \keyw{and}\;\, \ty{forest}_{\,\tau} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree}_{\,\tau},\:\, \ty{forest}_{\,\tau}) \\%[-.5\jot]
      \keyw{codatatype}\; \,\ty{llist}_{\,\tau} & \,=\, \const{LNil} \,\mid\, \const{LCons}(\const{lhead}{:}\; \tau,\; \const{ltail}{:}\;\ty{llist}_{\,\tau})
\end{aligned}
\]

@@

\paragraph{Properties.}
Datatypes and codatatypes share many basic properties pertaining to
constructors and selectors. All properties below are implicitly universally
quantified and range over all $i$, $j$, $\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}
  \\
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{k-1},y,x_{k+1},\ldots,x_{n_{i\negvthinspace j}}) \longrightarrow x_k \teq y
  \\
\text{Exhaustiveness:}\quad
  & \is{C}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \is{C}_{im_i}(x)
  \\
\text{Selection:}\quad
  & s_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) = x_k
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness ensures that ``no junk''
exists, whereas distinctness and injectivity ensure that ``no confusion''
arises. The result of selectors applied to the wrong constructor is left
completely unspecified.

@@

Datatypes are additionally characterized by an induction axiom schema:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\,i,j} \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\,k\,} \mathit{IH}_{i\negvthinspace j}^k[x_k]\bigr) \longrightarrow P_i[C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\,i} P_i[v_i]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{i\negvthinspace j}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$.
%
The axiom schema ensures that the interpretation of datatypes
is a smallest model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction rules out nonstandard models, which could contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.

For codatatypes, the dual notion is called coinduction. It depends on
witnesses $R$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\quad
\AXC{\strut$\begin{gathered}\textstyle \bigwedge_{\,i} R_i[v_i, w_i] \\[-\jot]\textstyle
\bigwedge_{\,i} \forall v\; w.\;\, R_i[v, w] \longrightarrow
  \bigwedge_{\,j} \const{d}_j(v) \teq \const{d}_j(w)
  \mathrel\land
  \const{d}_j(v) \longrightarrow \bigwedge_{\,k} \const{s}_{i\negvthinspace j}^{\,k}(v) \sim \const{s}_{i\negvthinspace j}^{\,k}(w)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\,i} v_i \teq w_i$}
\DP
\end{aligned}
\]
where $x \sim y$ denotes either $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
This axiom schema captures the idea that two values that yield the same
observations must be equal, where the observations are made by using the
selectors and discriminators.

@@

\paragraph{Semantics.}
Datatypes can be defined semantically as the initial model of the
selector--constructor equations. This is the view followed by Barrett et al.\
\cite{barrett-et-al-2010}. A drawback of this approach is that it does
not naturally account for selectors applied to wrong constructors. Barrett et
al.\ address this by parameterizing the construction by default values, but
these give rise to spurious equalities between unrelated terms---e.g.,
$\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$. This could be
corrected, but the added complexity seems to suggest that selectors are better
characterized axiomatically.

@@

Finally, some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\[\keyw{codatatype}\;\, \ty{a} \,=\, \const{A}(\ty{a})\]
This is a finite codatatype, of cardinality one, whose single value is
infinite: \,$\mu\vthinspace a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-\jot]
             \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$. We call such types \emph{corecursive singletons}. For the decision
procedure, it will be important to detect these types. A type may also be a
corecursive singleton only in some models. For example, if \ty{unit} is an
uninterpreted type, \ty{b} and \ty{c} will be singleton types precisely in the
models that interpret \ty{unit} as a singleton. Fortunately, given some
cardinalities for the ordinary types, it is easy to characterize this
degenerate case:

@@

\begin{lemma}
If the domain interpreting a corecursive codatatype is finite, it must be a
singleton. Furthermore, such a codatatype has necessarly a single constructor,
and all its arguments have types that are interpreted by singletons.
\end{lemma}

@@

@@

This work is part of a larger program that aims at enriching automatic
provers with more high-level features and reducing the gap between automatic and
interactive theorem proving. As future work, we want to try out CVC4 and its
(co)datatypes for higher-order model finding in Isabelle, as an
alternative to the counterexample generator Nitpick
\cite{blanchette-nipkow-2010}. We also see some opportunities to enrich
SMT solvers with recursive and corecursive functions. Finally, it might be
possible to go further in terms of supporting nested and mixed (co)recursion
and quantified formulas over (co)datatypes in solvers.

>>>>

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$). Symbols starting with an uppercase
letter (e.g.,
$\const{S}$, $\const{SCons}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed compactly as
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$,
which collapses to $\const{f} : \tau$ if $n = 0$.
For a term $t$, the notation $t : \tau$ indicates that it has type $\tau$.
Functions are invoked in the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
%
Finally, the notation $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$
and $\bigwedge_{\,i}\, \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$.

@@

We fix a signature $\Sig = (\Types, \Funcs)$. The types are partioned into
$\Types = \Data \mathrel{\uplus} \Codata \mathrel{\uplus} \Nondata$, where $\Data$ are the
\emph{datatypes}, $\Codata$ are the \emph{codatatypes}, and $\Nondata$ are the remaining
\emph{ordinary types}. The functions are partitioned into $\Funcs = \Ctr
\mathrel{\uplus} \Sel$, where $\Ctr$ are the \emph{constructors} and $\Sel$ are the
\emph{selectors}; there is no need to consider further function symbols
because they can be abstracted away as variables when combining theories.

\paragraph{Specifications.}
In an SMT problem, the signature is normally given by specifying the
uninterpreted types in any order, the (co)datatypes with their constructors
and selectors in groups of mutually (co)recursive groups of (co)datatypes, and
finally any other function symbols.
%
A (co)datatype specification consists of $l$~mutually recursive types which are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
In the presentation, we allow ourselves some metalevel parameterization
% at the metalevel
through subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}.

@@

For types with several constructors, it is customary to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. However,
it is not necessary to extend the signature:
The discriminator term $\const{d}_{i\negvthinspace j}(x)$ can be seen as an abbreviation for
$x = \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(x), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(x)\bigr)$.
%This will simplify the presentation. % of the decision procedure.

Here are a few examples of legal specifications of (co)datatype families:
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree}_{\,\tau} & \,=\, \const{Node}(\tau,\: \ty{forest}_{\,\tau}) \\[-\jot]
      \keyw{and}\;\, \ty{forest}_{\,\tau} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree}_{\,\tau},\:\, \ty{forest}_{\,\tau}) \\%[-.5\jot]
      \keyw{codatatype}\; \,\ty{llist}_{\,\tau} & \,=\, \const{LNil} \,\mid\, \const{LCons}(%\const{lhead}{:}\;
      \tau,\: %\const{ltail}{:}\;
      \ty{llist}_{\,\tau})
\end{aligned}
\]

@@

\paragraph{Properties.}
Datatypes and codatatypes share many basic properties pertaining to
constructors and selectors. All properties below are implicitly universally
quantified and range over all $i$, $j$, $\jPrime$, and $k$ within bounds:
% and to all possible splits of the
%$n$-ary constructor $\const{C}_{i\negvthinspace j}$'s argument list into $\bar x,
%y, \bar z$:
%
\[
\begin{aligned}[t]
\text{Distinctness:}\quad
  & %\forall \bar x, \bar y.\;\,
    \const{C}_{i\negvthinspace j}(\bar x) \tneq \const{C}_{i\negvthinspace \jPrime}(\bar y) \quad\text{if $j \not= \jPrime$}
  \\
\text{Injectivity:}\quad
  & %\forall x_1,\ldots,x_{n_{i\negvthinspace j}} y.\;\,
    \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}}) \teq \const{C}_{i\negvthinspace j}(x_1,\ldots,x_{k-1},y,x_{k+1},\ldots,x_{n_{i\negvthinspace j}}) \longrightarrow x_k \teq y
  \\
\text{Exhaustiveness:}\quad
  & \is{C}_{i1}(x) \mathrel\lor \cdots \mathrel\lor \is{C}_{im_i}(x)
  \\
\text{Selection:}\quad
  & s_{i\negvthinspace j}^{\,k}(\const{C}_{i\negvthinspace j}(x_1,\ldots,x_{n_{i\negvthinspace j}})) = x_k
\end{aligned}
\]
%
Expressed in the algebraic jargon, exhaustiveness helps ensure that ``no
junk'' exists, whereas distinctness and injectivity guarantee that ``no
confusion'' can arise. The result of selectors applied to the wrong
constructor is left completely unspecified.

@@

Datatypes are additionally characterized by an induction axiom schema:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\,i,j}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\,k}\, \mathit{IH}_{i\negvthinspace j}^k[x_k]\bigr) \longrightarrow P_i[C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\,i}\, P_i[v_i]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{i\negvthinspace j}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$.
%
The axiom schema ensures that the interpretation of datatypes
is a standard model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction rules out nonstandard models, which could contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.

For codatatypes, the dual notion is called coinduction. It depends on
witnesses $R_i$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Coinduction:}\kern.8em %%% TYPESETTING: should be \quad
\AXC{\strut$\begin{gathered}\textstyle \bigwedge_{\,i}\, R_i[v_i, w_i] \\[-\jot]\textstyle
\bigwedge_{\,i}\, \forall v\; w.\;\, R_i[v, w] \longrightarrow
  \bigwedge_{\,j}\, \const{d}_j(v) \teq \const{d}_j(w)
  \mathrel\land
  \const{d}_j(v) \longrightarrow \bigwedge_{\,k}\, \const{s}_{i\negvthinspace j}^{\,k}(v) \sim \const{s}_{i\negvthinspace j}^{\,k}(w)
\end{gathered}$}
\UIC{\strut$\bigwedge_{\,i}\, v_i \teq w_i$}
\DP
\end{aligned}
\]
where $x \sim y$ denotes either $R_{\iPrime}[x, y]$ if there exists some
$\iPrime$ such that the formula is type-correct or $x \teq y$ otherwise.
This axiom schema captures the idea that two values that yield the same
observations must be equal, where the observations are made by using the
selectors and discriminators.

@@

\paragraph{Semantics.}
Datatypes can be defined semantically as the initial model of the
selector--constructor equations. This is the view followed by Barrett et al.\
\cite{barrett-et-al-2010}. A drawback of this approach is that it does
not naturally account for selectors applied to wrong constructors. Barrett et
al.\ address this by parameterizing the construction by default values, but
this gives rise to spurious equalities between unrelated terms---e.g.,
$\const{s}_{11}(\const{C}_2) \teq \const{s}_{11}(\const{C}_3)$. This flaw
could be corrected, but the added complexity seems to suggest that selectors
are better characterized axiomatically.

@@

Finally, some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\[\keyw{codatatype}\;\, \ty{a} \,=\, \const{A}(\ty{a})\]
This is a finite codatatype, of cardinality one, whose single value is
infinite: \,$\mu\vthinspace a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-\jot]
             \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$, as well as $\ty{stream}_{\,\ty{unit}\,}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
important to detect these types, even if they occur rarely in practice. A type
may also be a corecursive singleton only in some models. For example, if
\ty{unit} is an uninterpreted type, \ty{b} and \ty{c} will be singleton types
precisely in the models that interpret \ty{unit} as a singleton. Fortunately,
given some cardinalities for the ordinary types, it is easy to characterize
this degenerate case:

@@

\begin{lemma}
Let $\delta$ be a corecursive codatatype. If the domain interpreting $\delta$
is finite, it must be a singleton. Furthermore, $\delta$ has necessarily a
single constructor, whose arguments have types that are interpreted by
singletons.
\end{lemma}

@@

  * to evaluate procedure, we generated problems from Isabelle theories using
    Sledgehammer

  * chose a selection of theory files:
    * pretty much every theory file from Isabelle distribution and AFP that
      defines a codatatype
    * also included two private theories by Andreas Lochbihler and Peter Gammie
    * (codatatypes being a recent addition to Isabelle \cite{itp},
       they are not yet widely used)
    * and included a number of theories on datatypes
    * aimed at a good mixture between linear structures (lists, streams) and
      tree structures
    * the theory files were chosen before running the experiments

  * for each goal, select 256 lemmas as axioms, before monomorphization
    (after monomorphization, typically more formulas due to duplicates),
    and encode with the goal

  * then ran CVC4 for XXX seconds on the StarExec cluster \cite{xxx}

{\footnotesize
\begin{verbatim}
For the evaluation of Section 5, I'm playing with various ideas. I certainly
would like to keep it short and not waste too much of our time on it. We
discussed earlier having (1) hand-crafted problems and (2) actual benchmarks
(like those you already have). I suggest we drop (1); I never did hand-crafted
before and this always looks suspicious anyway. For (2), I see three or four
possible benchmark sets:

1. without (co)datatype theories
2. with datatypes
(3. with codatatypes)
4. with both

The benchmark problems would be identical except for the declarations at the
top ("declare-sort" and "declare-fun" vs. "declare-(co)datatype").
Irrespective of whether interpreted (co)datatypes are used, Sledgehammer will
typically include some (co)datatype axioms (i.e. tautologies if (co)datatypes
are enabled). Hence the tests are a good indication of whether native
(co)datatypes are useful to Sledgehammer. I assume the results will follow
common sense, i.e. 4 > {3, 2} > 1

As for the actual Isabelle theory files to use, for codatatypes there is not
much question: We just use all I could get my hands on. But perhaps we should
have a second benchmark set, ideally something a bit standard like, say, the
Judgment Day (JD) suite. Or I can make some arbitrary a priori (before
evaluation) decisions, but that always looks a bit fishy unless it can be
justified easily. The trouble is, we have about 1000 datatypes lying around in
Isabelle and the Archive of Formal Proofs (AFP). These would have no
codatatypes in them, so we could only do 1 and 2 of the evaluation. We could
at least try the "list" library, which is extremely developed and would have
1600 goals already, and perhaps a few really basic types ("option", "string"),
or perhaps red-black trees (about 450 goals).

So we can show the success rates, in a table like this:

                 Codatatype-oriented suite      Datatype (or JD) suite
                      (2345 goals)                   (1240 goals)

   w/o (co)data        1255                            699
   w data              1268                            712
   w codata            1270
   w both              1273

I can write the usual warning that an increase of a couple of percentage
points in success rate in a general tool like Sledgehammer and general
benchmark suites means a lot etc. Putting the datatype evaluations (i.e. 2 and
4) in there offers the advantage that it will give some measure of how much to
expect from the codatatypes. I.e. if the success rate doesn't go up by 50% for
datatypes, nobody will expect this to happen for codatatypes. A priori, I
would expect a similar gain, but I'm looking forward to see the experimental
data.

In addition to success rate (or number of solved goals), we should have other
statistics, like how often do we get proofs that contain the cyclicity and
bisimilarity rules. And finally, we look in more detail at one or two such
proofs, by presenting it and explaining it (and, before that, understanding
it).
\end{verbatim}}

@@

This work is part of a larger program that aims at enriching automatic
provers with high-level features and at reducing the gap between automatic and
interactive theorem proving. As future work, we want to try out CVC4 and its
(co)datatypes for higher-order model finding in Isabelle, as an
alternative to the counterexample generator Nitpick
\cite{blanchette-nipkow-2010}. We also see some opportunities to enrich
SMT solvers with recursive and corecursive functions. Finally, it might be
possible to go further in terms of supporting nested and mixed (co)recursion
and quantified formulas over (co)datatypes in solvers.
